The role of advection versus diffusion in intraventricular transport

### 2025-03-03 paper/main.tex
Spell-check, move references up, and comment out boilerplate stuff.

To date, it remains poorly understood how cilia contribute to the
overall fluid flow and solute transport within the brain.
These questions have been challenging to tackle due to technical
limitations. Notably, cilia are microscopic and beat at high
frequencies ranging from 10 to 30 Hz, while other fluid dynamic
processes in the brain are slower and span larger volumes, making
cilia dynamics challenging to monitor in vivo. Until now, most studies have
analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF flow
and the accompanying transport. Similarly, while MRI technology can measure CSF movement
in humans, it cannot detect the microscopic movements of cilia to
understand their impact in the large human brain. Consequently, we
lack a comprehensive understanding of how cilia contribute to CSF
motion and solute transport, and whether cilia dysfunction will
alter CSF composition and lead to brain disorders. This is particularly
important since ventricular enlargement, known as hydrocephalus,
is observed in animal models and humans with motile cilia deficiency,
indiciating that the cilia play a crucial role for healthy brain functioning.

@@

\subsection{Modification of ventricles geometry impacts protein distribution}
Variations in ventricular morphology impact CSF flow patterns.
To assess how ventricular morphologies influence solute distribution,
we consider four alternative brain ventricles meshes:
one where the duct connecting the anterior and middle ventricles is constricted
(66 \% reduction in cross-sectional area);
one where the duct connecting the middle and posterior ventricles is constricted
(33 \% reduction);
one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
and the combination of the above three modifications (\Crefpanels{fig:fig6}{a}). In these
four modified geometries the volume is reduced by 3.9\%, 4.1\%, 1.4\% and 9.4\%, respectively,
comapred to the original geometry.
We simulate CSF flow in the modified geometries with the baseline flow model
and use the resulting velocity field to simulate transport of Dendra2
(\Crefpanels{fig:fig6}{b}).

@@

Using the simulated CSF flow, we modeled solute transport by an advection-diffusion equation.
The transport results compared well with experimental data performed in vivo in zebrafish.
Then, by using a computational model as well as zebrafish mutants with paralyzed cilia,
we identified that even though advection by cilia is dominant,
diffusion plays a significant role in protein transport.
Finally, we reported how local cilia paralysis and ventricular geometries
affect solute distribution. Altogether, we present not only new findings on the role of
diffusion and advection in intraventriuclar transport, but also provide a novel framework to
study CSF flow and transport in complex geometries.

@@

\subsection{Effects of cilia motility and ventricular morphology}
To gain further insights into how CSF flow and intraventricular transport depend on cilia
motility and ventricular morphology, we manipulated certain
parameters in our models. First, we investigated how complete loss of cilia motility
impacted the transport dynamics. Simulations with a cardiac-only/no-cilia lead to
slower transport when compared to the baseline model that included cilia. Especially
the middle-ventricular region downstream of CSF flow (ROI 2) and the anterior ventricle (ROI 5)
experienced substantially increased times-to-threshold. We also observed
slower dynamics in the rest of the middle ventricle (ROIs 3 and 4), whereas the
posterior ventricle (ROI 6) was unaffected by the loss
of cilia motility, indicating that transport from the middle to the posterior ventricle is
purely driven by diffusion. We also performed photoconversion experiments on cilia mutant
zebrafish larva (\emph{smh}), observing statistically signifcant slower transport in 
ROIs 2 and 4. The results provide not only confidence in that our computational model
well represents the physics, but also valuable quantification of the impact of cilia motility
on intraventricular transport. Interestingly, a previous simulation study of reduced cilia motility
also observed a drastic reduction in CSF mixing in specific regions of the brain
ventricles~\cite{Yoshida2022EffectVentricles}, similar to our results. Since impaired brain 
clearance of tracers has been demonstrated in idiopathic normal pressure hydrocephalus
patients~\cite{Eide2020MagneticHydrocephalus}, cilia-enhanced mixing in the brain might be
fundamental to physiological transport within the brain.

@@

With an interest in assessing the impact of ventricular morphology on solute transprot,
we simulated transport on modified ventricular geometries.
Notably, shrinking ducts reduced interventricular solute movement, while shrinking
the middle ventricle reduced diffusive transport upstream of the flow field.
We previously identified large inter-individual variation in ventricular size and
CSF velocities between larva at given developmental stages and throughout
development~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain, Jeong2024TheZebrafish}.
Our results suggest that these two parameters, CSF velocities and ventricular size,
may well co-vary and explain the variation in the two.

@@

\subsection{The role of motile cilia in intraventricular flow and transport}
Our work has studied CSF flow and solute distribution
in a relatively small and simple brain model, which is devoid of
active CSF secretion and CSF absorption. Even though there is
a clear conservation of the ventricular system across vertebrates,
there are also important considerations to take. These include (but are not limitied to)
ventricular size and morphology, properties of ciliated cells,
animal posture, and magnitude of velocities. For instance, a large
volume would require an extensive amount of time for solutes to distribute
if based solely on diffusion. Besides, the presence of multiple cilia
instead of a single cilium per cell -- the former can be found in
mammalian ependymal cells -- will generate stronger velocities.
Hence advection may be more dominant in those situations.

@@

% Acknowledgements
\medskip
\textbf{Acknowledgements} \par 
Thanks are extended to J\o rgen Dokken for valuable model implementation discussions.
Halvor Herlyng acknowledges support from the national infrastructure for computational
science in Norway, Sigma2, via grant \#NN8049K.
This work was funded by the Research Council of Norway (NJY FRIPRO grants 314189 and 353348).

\medskip
\textbf{Data Availability Statement} \par
The data and implementations presented in this paper are licensed under \fixme{LICENSE}
and openly avalaible at \fixme{REPO-LINK}.

@@

%%--------------APPENDIX--------------%%
\appendix
\section{Supplementary Theory}
\subsection{Weak formulation of the Stokes equations}\label{subsec:appendixA1}
We consider the time-dependent Stokes equations
\begin{subequations}
\begin{alignat}{2}
   \rho\pdifft{\uu} - \nabla \cdot \bsig(\uu, p)
   &= \mathbf{0} \quad &&\mathrm{in} \ \Omega, \label{eq:BDM_stokes_mom}\\
  \nabla \cdot \uu &= 0 \quad &&\mathrm{in} \ \Omega, \label{eq:BDM_stokes_div} 
\end{alignat}
\label{eq:appendix_stokes_eqs}% 
\end{subequations}%
in the domain $\Omega$ with boundary $\Gamma$, where
$\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$ is the viscous
stress tensor, in which $\beps(\uu) = \frac{1}{2}(\nabla\uu + (\nabla\uu)^T)$
is the symmetric gradient. Furthermore, $\uu\in\VV$ is the fluid velocity
and $p\in Q$ is the fluid pressure, and $\rho$ and $\mu$ are the fluid density
and dynamic viscosity, respectively. Unless otherwise stated, we let $\bsig = \bsig(\uu, p)$.

@@

\textbf{References}\\
1	((Journal articles)) a) A. B. Author 1, C. D. Author 2, Adv. Mater. 2006, 18, 1; b) A. Author 1, B. Author 2, Adv. Funct. Mater. 2006, 16, 1.\\
2	((Work accepted)) A. B. Author 1, C. D. Author 2, Macromol. Rapid Commun., DOI: 10.1002/marc.DOI.\\
3	((Books)) H. R. Allcock, Introduction to Materials Chemistry, Wiley, Hoboken, NJ, USA 2008.\\
4	((Edited books or proceedings volumes)) J. W. Grate, G. C. Frye, in Sensors Update, Vol. 2 (Eds: H. Baltes, W. Göpel, J. Hesse), Wiley-VCH, Weinheim, Germany 1996, Ch. 2.\\
5	((Presentation at a conference, proceeding not published)) Author, presented at Abbrev. Conf. Title, Location of Conference, Date of Conference ((Month, Year)).\\
6	((Thesis)) Author, Degree Thesis, University (location if not obvious), Month, Year.\\
7	((Patents)) a) A. B. Author 1, C. D. Author 2 (Company), Country Patent Number, Year; b) W. Lehmann, H. Rinke (Bayer AG) Ger. 838217, 1952.\\
8	((Website)) Author, Short description or title, URL, accessed: Month, Year.\\
9	…((Please include all authors, and do not use “et al.”))\\

@@

\begin{figure}
\textbf{Table of Contents}\\
\medskip
  \includegraphics{toc-image.png}
  \medskip
  \caption*{ToC Entry}
\end{figure}

>>>>

To date, it remains poorly understood how cilia contribute to the
overall fluid flow and solute transport within the brain.
These questions have been challenging to tackle due to technical
limitations. Notably, cilia are microscopic and beat at high
frequencies ranging from 10 to 30 Hz, while other fluid dynamic
processes in the brain are slower and span larger volumes, making
cilia dynamics challenging to monitor in vivo. Until now, most studies have
analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF flow
and the accompanying transport. Similarly, while MRI technology can measure CSF movement
in humans, it cannot detect the microscopic movements of cilia to
understand their impact in the large human brain. Consequently, we
lack a comprehensive understanding of how cilia contribute to CSF
motion and solute transport, and whether cilia dysfunction will
alter CSF composition and lead to brain disorders. This is particularly
important since ventricular enlargement, known as hydrocephalus,
is observed in animal models and humans with motile cilia deficiency,
indicating that the cilia play a crucial role for healthy brain functioning.

@@

\subsection{Modification of ventricles geometry impacts protein distribution}
Variations in ventricular morphology impact CSF flow patterns.
To assess how ventricular morphologies influence solute distribution,
we consider four alternative brain ventricles meshes:
one where the duct connecting the anterior and middle ventricles is constricted
(66 \% reduction in cross-sectional area);
one where the duct connecting the middle and posterior ventricles is constricted
(33 \% reduction);
one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
and the combination of the above three modifications (\Crefpanels{fig:fig6}{a}). In these
four modified geometries the volume is reduced by 3.9\%, 4.1\%, 1.4\% and 9.4\%, respectively,
compared to the original geometry.
We simulate CSF flow in the modified geometries with the baseline flow model
and use the resulting velocity field to simulate transport of Dendra2
(\Crefpanels{fig:fig6}{b}).

@@

Using the simulated CSF flow, we modeled solute transport by an advection-diffusion equation.
The transport results compared well with experimental data performed in vivo in zebrafish.
Then, by using a computational model as well as zebrafish mutants with paralyzed cilia,
we identified that even though advection by cilia is dominant,
diffusion plays a significant role in protein transport.
Finally, we reported how local cilia paralysis and ventricular geometries
affect solute distribution. Altogether, we present not only new findings on the role of
diffusion and advection in intraventricular transport, but also provide a novel framework to
study CSF flow and transport in complex geometries.

@@

\subsection{Effects of cilia motility and ventricular morphology}
To gain further insights into how CSF flow and intraventricular transport depend on cilia
motility and ventricular morphology, we manipulated certain
parameters in our models. First, we investigated how complete loss of cilia motility
impacted the transport dynamics. Simulations with a cardiac-only/no-cilia lead to
slower transport when compared to the baseline model that included cilia. Especially
the middle-ventricular region downstream of CSF flow (ROI 2) and the anterior ventricle (ROI 5)
experienced substantially increased times-to-threshold. We also observed
slower dynamics in the rest of the middle ventricle (ROIs 3 and 4), whereas the
posterior ventricle (ROI 6) was unaffected by the loss
of cilia motility, indicating that transport from the middle to the posterior ventricle is
purely driven by diffusion. We also performed photoconversion experiments on cilia mutant
zebrafish larva (\emph{smh}), observing statistically significant slower transport in 
ROIs 2 and 4. The results provide not only confidence in that our computational model
well represents the physics, but also valuable quantification of the impact of cilia motility
on intraventricular transport. Interestingly, a previous simulation study of reduced cilia motility
also observed a drastic reduction in CSF mixing in specific regions of the brain
ventricles~\cite{Yoshida2022EffectVentricles}, similar to our results. Since impaired brain 
clearance of tracers has been demonstrated in idiopathic normal pressure hydrocephalus
patients~\cite{Eide2020MagneticHydrocephalus}, cilia-enhanced mixing in the brain might be
fundamental to physiological transport within the brain.

@@

With an interest in assessing the impact of ventricular morphology on solute transport,
we simulated transport on modified ventricular geometries.
Notably, shrinking ducts reduced interventricular solute movement, while shrinking
the middle ventricle reduced diffusive transport upstream of the flow field.
We previously identified large inter-individual variation in ventricular size and
CSF velocities between larva at given developmental stages and throughout
development~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain, Jeong2024TheZebrafish}.
Our results suggest that these two parameters, CSF velocities and ventricular size,
may well co-vary and explain the variation in the two.

@@

\subsection{The role of motile cilia in intraventricular flow and transport}
Our work has studied CSF flow and solute distribution
in a relatively small and simple brain model, which is devoid of
active CSF secretion and CSF absorption. Even though there is
a clear conservation of the ventricular system across vertebrates,
there are also important considerations to take. These include (but are not limited to)
ventricular size and morphology, properties of ciliated cells,
animal posture, and magnitude of velocities. For instance, a large
volume would require an extensive amount of time for solutes to distribute
if based solely on diffusion. Besides, the presence of multiple cilia
instead of a single cilium per cell -- the former can be found in
mammalian ependymal cells -- will generate stronger velocities.
Hence advection may be more dominant in those situations.

@@

% Acknowledgements
\medskip
\textbf{Acknowledgments} \par 
Thanks are extended to J\o rgen Dokken for valuable model implementation discussions.
Halvor Herlyng acknowledges support from the national infrastructure for computational
science in Norway, Sigma2, via grant \#NN8049K.
This work was funded by the Research Council of Norway (NJY FRIPRO grants 314189 and 353348).

\medskip
\textbf{Data Availability Statement} \par
The data and implementations presented in this paper are licensed under \fixme{LICENSE}
and openly available at \fixme{REPO-LINK}.

@@

\newpage 
%%--------------APPENDIX--------------%%
\appendix
\section{Supplementary Theory}
\subsection{Weak formulation of the Stokes equations}\label{subsec:appendixA1}
We consider the time-dependent Stokes equations
\begin{subequations}
\begin{alignat}{2}
   \rho\pdifft{\uu} - \nabla \cdot \bsig(\uu, p)
   &= \mathbf{0} \quad &&\mathrm{in} \ \Omega, \label{eq:BDM_stokes_mom}\\
  \nabla \cdot \uu &= 0 \quad &&\mathrm{in} \ \Omega, \label{eq:BDM_stokes_div} 
\end{alignat}
\label{eq:appendix_stokes_eqs}% 
\end{subequations}%
in the domain $\Omega$ with boundary $\Gamma$, where
$\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$ is the viscous
stress tensor, in which $\beps(\uu) = \frac{1}{2}(\nabla\uu + (\nabla\uu)^T)$
is the symmetric gradient. Furthermore, $\uu\in\VV$ is the fluid velocity
and $p\in Q$ is the fluid pressure, and $\rho$ and $\mu$ are the fluid density
and dynamic viscosity, respectively. Unless otherwise stated, we let $\bsig = \bsig(\uu, p)$.

@@

%% \textbf{References}\\
%% 1	((Journal articles)) a) A. B. Author 1, C. D. Author 2, Adv. Mater. 2006, 18, 1; b) A. Author 1, B. Author 2, Adv. Funct. Mater. 2006, 16, 1.\\
%% 2	((Work accepted)) A. B. Author 1, C. D. Author 2, Macromol. Rapid Commun., DOI: 10.1002/marc.DOI.\\
%% 3	((Books)) H. R. Allcock, Introduction to Materials Chemistry, Wiley, Hoboken, NJ, USA 2008.\\
%% 4	((Edited books or proceedings volumes)) J. W. Grate, G. C. Frye, in Sensors Update, Vol. 2 (Eds: H. Baltes, W. Göpel, J. Hesse), Wiley-VCH, Weinheim, Germany 1996, Ch. 2.\\
%% 5	((Presentation at a conference, proceeding not published)) Author, presented at Abbrev. Conf. Title, Location of Conference, Date of Conference ((Month, Year)).\\
%% 6	((Thesis)) Author, Degree Thesis, University (location if not obvious), Month, Year.\\
%% 7	((Patents)) a) A. B. Author 1, C. D. Author 2 (Company), Country Patent Number, Year; b) W. Lehmann, H. Rinke (Bayer AG) Ger. 838217, 1952.\\
%% 8	((Website)) Author, Short description or title, URL, accessed: Month, Year.\\
%% 9	…((Please include all authors, and do not use “et al.”))\\

@@

%% \begin{figure}
%% \textbf{Table of Contents}\\
%% \medskip
%%   \includegraphics{toc-image.png}
%%   \medskip
%%   \caption*{ToC Entry}
%% \end{figure}
